Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan

### 2020-07-29 analysis/paper/paper.Rmd
adjust number in regression plot, and make boxplot for cv of size vs shape

Indigenous societies' responses to colonial contact ranges from passive acceptance to active negotiation with the colonists, and accommodation or resistance of foreign intrusion, which can be identified through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000]. In this paper we investigate the archaeology of a pericolonial situation at Kiwulan (ca. AD 1350-1850), a large multi-component archaeological site in Yilan County, northeastern Taiwan [@Chen2007], to identify the indirect impacts of colonial settler activity on local Indigenous societies. Yilan is an ideal context to study peripheral colonial influences because the Indigenous communities were isolated by geographical barriers, limiting the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan. Kiwulan is situated on a hill near a riverside at the northern margin of Yilan County, which is characterized by a triangular alluvial plain facing east toward the Pacific with high mountains on three other sides. 
This research investigates if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, which may indicate social changes in local Indigenous societies. We predict that competition within the Indigenous community at Kiwulan for foreign resources and trade partnerships with European or Chinese colonizers may have led to the emergence of craft specialization caused by greater economic and social control of ceramic production by a small group of individuals. Using standardization in ceramic shapes as a proxy for craft specialization, we ask: Did colonial trade impact the shape of locally-produced Indigenous pottery vessels? Did pottery shape become more homogeneous and standardized after foreign contacts with European colonizers or Chinese immigrants?

@@

Geometric morphometrics (GMM) differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to quantify and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric methods: landmark and outline approaches [@Adams2004]. Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points. Generalized Procrustes analysis (GPA) is used to superimpose landmark data on a common coordinate system by translating, rotating, and scaling [@Bookstein1991]. After the procedure of GPA, superimposed landmark coordinates become shape variables that allow further statistical analyses [@Slice2007]. Landmark-based morphometrics have been widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with distinct components [@Selden2019; @Topi2017]. With univariate or multivariate statistics, the GMM applications can answer research questions related to lithic typological and technological change [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], and hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019]. A common procedure is using dimensional reduction techniques, such as Principal Components Analysis or Canonical Variate Analysis, to capture the key features that represent overall shape. Visualization of the reduced data enables the identification of groups, followed by statistical tests to robustly distinguish them.

Key questions in archaeological shape analysis normally involve measuring shape standardization over time, or between geographical areas. Standardization is often investigated using multivariate analysis of shape variables computed from landmark data, or coefficients of variation on associated metric data, especially for lithic assemblages. For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to increased maintenance of finished points. @Buchanan2018 studied metric data and GMM to identify a more uniform base-shape of Folsom points than Clovis points across the western US [@Buchanan2019]. Also using metric data and GMM, @Smith2016 found standardized bases of fluted point in Alaska and northern Yukon that might indicate a risk management strategy to ensure the ease of replacement during long-distance travel. Other factors, such as low levels of cultural innovation in a small group, could also lead to an increase in standardization of point shapes [@Okumura2014]. For ceramics, @Topi2017 identified standardized vessel shapes from the Casas Grandes culture in northwest Mexico using coefficients of variation for the positions of semi-landmarks across shape groups. They suggested standardization might hint the presence of specialized producers, reflecting social complexity. Another way to explore standardization is pairwise testing of variations in morphological disparity among groups of shape variables based on the calculation of distances between groups in morphospace they occupy [@Wills2001]. In this manner, @Selden2018ceramic2 examined Caddo ceramics using semi-landmark approaches in northeast Texas and found a gradual increase in shape standardization over time that provides a basis for further discussion of craft specialization or group identity [@Selden2019]. Similar applications to lithic assemblages suggests that the Gahagan bifaces from the central Texas exhibit less size standardization than those from the southern Caddo area, indicative of different uses  [@Selden2020]. Other applications, such as the study of cranial deformation, demonstrate that landmark approaches with multivariate analyses of shape variances are an effective method to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].

@@

```{r PCs-distribution, fig.cap= "The distribution of PC scores by phases"}

# function for normalized data 
# normalize <- function(x) (x-min(x))/(max(x)-min(x)) # between 0-1
normalize <- function(x) (1 + ((x-min(x))* 2))/(max(x)-min(x))

@@

ggplot(pc_plot,
       aes(phase,
           pc)) +
  geom_violin() + 
  ggforce::geom_sina() +
  xlab("") +
  facet_wrap(~ pcn, ncol = 1) +
  geom_text(data = cv_pcs_all,
            mapping = aes(x = c(2.1, 1.1, 3.1, 2.1, 1.1, 3.1, 2.1, 1.1, 3.1), 
                          y = 2.1, 
                          label = cv_label),
            size = 3,
            color= "red",
            hjust = -0.1,
            vjust= -1)+
  theme_minimal()

@@

pc2_chinese <-
  pc2 %>% 
  filter(y == "Chinese") %>% 
  mutate(cv = round(sd(x)/ mean(x), digits = 2)*100)

@@

# can pc1 predict size by period?
pc1_size_reg <- 
  ggscatter(jpgs_imported_coo_size_pcs_metric, 
            x = "Diameter1_Body", 
            y = "pc1",
            add = "reg.line",
            size = 1) +
    facet_wrap(~ fct_rev(y)) +
    stat_cor(label.y = 1.25, label.sep = "\n") +
    #stat_regline_equation(label.y = 1) +
    labs(x = "", y = "PC1") +
    theme(axis.text = element_text(size = 8)) +
    #scale_x_continuous(breaks = c(75, 200)) +
    scale_y_continuous(limits = c(0, 1.5),
                       breaks = c(seq(0, 1, 0.5)),
                       labels = c(seq(0, 1, 0.5)))

# can pc2 predict size by period?
pc2_size_reg <- 
  ggscatter(jpgs_imported_coo_size_pcs_metric, 
            x = "Diameter1_Body", 
            y = "pc2",
            add = "reg.line",
            size = 1) +
    facet_wrap(~ fct_rev(y)) +
    stat_cor(label.y = 1.25, label.sep = "\n") +
    #stat_regline_equation(label.y = 1) +
    labs(x = "body diameter (mm)", y = "PC2") +
    theme(axis.text = element_text(size = 8)) +
    #scale_x_continuous(breaks = c(75, 200)) +
    scale_y_continuous(limits = c(0, 1.5),
                       breaks = c(seq(0, 1, 0.5)),
                       labels = c(seq(0, 1, 0.5)))

@@

```{r boxplot-pc1-ave-size}
size_pcs_cv_join <-
  jpgs_imported_coo_size_pcs_metric %>% 
  left_join(cv_pcs_all, by = c("y", "phase"))

>>>>

Indigenous societies' responses to colonial contact ranges from passive acceptance to active negotiation with the colonists, and accommodation or resistance of foreign intrusion, which can be identified through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000]. In this paper we investigate the archaeology of a pericolonial situation at Kiwulan (ca. AD 1350-1850), a large multi-component archaeological site in Yilan County, northeastern Taiwan [@Chen2007], to identify the indirect impacts of colonial settler activity on local Indigenous societies. Yilan is an ideal context to study peripheral colonial influences because the Indigenous communities were isolated by geographical barriers, limiting the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan. Kiwulan is situated on a hill near a riverside at the northern margin of Yilan County, which is characterized by a triangular alluvial plain facing east toward the Pacific with high mountains on three other sides.

This research investigates if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, which may indicate social changes in local Indigenous societies. We predict that competition within the Indigenous community at Kiwulan for foreign resources and trade partnerships with European or Chinese colonizers may have led to the emergence of craft specialization caused by greater economic and social control of ceramic production by a small group of individuals. Using standardization in ceramic shapes as a proxy for craft specialization, we ask: Did colonial trade impact the shape of locally-produced Indigenous pottery vessels? Did pottery shape become more homogeneous and standardized after foreign contacts with European colonizers or Chinese immigrants?

@@

Geometric morphometrics (GMM) differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to quantify and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric methods: landmark and outline approaches [@Adams2004]. Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points. Generalized Procrustes analysis (GPA) is used to superimpose landmark data on a common coordinate system by translating, rotating, and scaling [@Bookstein1991]. After the procedure of GPA, superimposed landmark coordinates become shape variables that allow further statistical analyses [@Slice2007]. A common procedure is using dimensional reduction techniques, such as Principal Components Analysis or Canonical Variate Analysis, to capture the key features that represent overall shape. Visualization of the reduced data enables the identification of groups, followed by statistical tests to robustly distinguish them. Landmark-based morphometrics have been widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with distinct components [@Selden2019; @Topi2017]. GMM applications can answer research questions related to lithic typological and technological change [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], or hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019].

Key questions in archaeological shape analysis normally involve measuring shape standardization over time, or between geographical areas. Standardization is often investigated using multivariate analysis of shape variables computed from landmark data, or coefficients of variation on associated metric data, especially for lithic assemblages. For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to increased maintenance of finished points. @Buchanan2018 studied metric data and GMM to identify a more uniform base-shape of Folsom points than Clovis points across the western US [@Buchanan2019]. Also using metric data and GMM, @Smith2016 found standardized bases of fluted point in Alaska and northern Yukon that might indicate a risk management strategy to ensure the ease of replacement during long-distance travel. Other factors, such as low levels of cultural innovation in a small group, could also lead to an increase in standardization of point shapes [@Okumura2014].

For ceramics, @Topi2017 identified standardized vessel shapes from the Casas Grandes culture in northwest Mexico using coefficients of variation for the positions of semi-landmarks across shape groups. They suggested standardization might hint the presence of specialized producers, reflecting social complexity. Another way to explore standardization is pairwise testing of variations in morphological disparity among groups of shape variables based on the calculation of distances between groups in morphospace they occupy [@Wills2001]. In this manner, @Selden2018ceramic2 examined Caddo ceramics using semi-landmark approaches in northeast Texas and found a gradual increase in shape standardization over time that provides a basis for further discussion of craft specialization or group identity [@Selden2019]. Similar applications to lithic assemblages suggests that the Gahagan bifaces from the central Texas exhibit less size standardization than those from the southern Caddo area, indicative of different uses  [@Selden2020]. Other applications, such as the study of cranial deformation, demonstrate that landmark approaches with multivariate analyses of shape variances are an effective method to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].

@@

```{r PCs-distribution, fig.cap= "The distribution of mormalized PC scores by phases with CV values (%)"}

# function for normalized data 
# normalize <- function(x) (x-min(x))/(max(x)-min(x)) # between 0-1
normalize <- function(x) (1 + (x-min(x)))/(max(x)-min(x))

@@

ggplot(pc_plot,
       aes(phase,
           pc)) +
  geom_violin() + 
  ggforce::geom_sina() +
  xlab("") +
  facet_wrap(~ pcn, ncol = 1) +
  geom_text(data = cv_pcs_all,
            mapping = aes(x = c(2.1, 1.1, 3.1, 2.1, 1.1, 3.1, 2.1, 1.1, 3.1), 
                          y = 1.8, 
                          label = cv_label),
            size = 3,
            color= "red",
            hjust = -0.1,
            vjust= -1)+
  theme_minimal()

@@

@@

# can pc1 predict size by period?
pc1_size_reg <- 
  ggscatter(jpgs_imported_coo_size_pcs_metric, 
            x = "Diameter1_Body", 
            y = "pc1",
            add = "reg.line",
            size = 1) +
    facet_wrap(~ fct_rev(y)) +
    stat_cor(label.y = 2.25, label.sep = "\n") +
    #stat_regline_equation(label.y = 1) +
    labs(x = "", y = "PC1") +
    theme(axis.text = element_text(size = 8)) +
    #scale_x_continuous(breaks = c(75, 200)) +
    scale_y_continuous(limits = c(1, 2.5),
                       breaks = c(seq(1, 2.5, 0.5)),
                       labels = c(seq(1, 2.5, 0.5)))

# can pc2 predict size by period?
pc2_size_reg <- 
  ggscatter(jpgs_imported_coo_size_pcs_metric, 
            x = "Diameter1_Body", 
            y = "pc2",
            add = "reg.line",
            size = 1) +
    facet_wrap(~ fct_rev(y)) +
    stat_cor(label.y = 2.25, label.sep = "\n") +
    #stat_regline_equation(label.y = 1) +
    labs(x = "body diameter (mm)", y = "PC2") +
    theme(axis.text = element_text(size = 8)) +
    #scale_x_continuous(breaks = c(75, 200)) +
    scale_y_continuous(limits = c(1, 2.5),
                       breaks = c(seq(1, 2.5, 0.5)),
                       labels = c(seq(1, 2.5, 0.5)))

@@

```{r boxplot-pc1-ave-size}
size_pcs_cv_join <-
  jpgs_imported_coo_size_pcs_metric %>% 
  group_by(y) %>% 
  mutate(Diameter1_Body_ave = round(mean(Diameter1_Body), 2),
         cv_body_diameter = round(sd(Diameter1_Body)/mean(Diameter1_Body), 
                                  digits = 2)*100) %>%
  mutate(phase = as.factor(y)) %>% 
  left_join(cv_pcs_all) %>% 
  left_join(pc1_df_with_filename)
  #distinct(phase, cv_pc1, cv_pc2, cv_pc3, Diameter1_Body_ave, cv_body_diameter)

boxplot_ave_size_cv_pc <-
  size_pcs_cv_join %>% 
  ggplot(aes(cv_body_diameter, cv_pc1, shape = phase)) +
  geom_point() +
  coord_flip() +
  labs(x = "CV of Body diameter (%)",
       y = "CV of PC1 (%)")
```
